---
title: "ST 557 Final Project"
author: "Emmanuel Asante (933274210)"
date: "November 28, 2017"
output: pdf_document
---

```{r,include=F,echo=F,warning=F,message=F}
library(mvtnorm)
library(cluster)
library(ICSNP)
library(plotrix)
library(graphics)
library(tidyverse)
library(purrr)
library(dplyr)
library(mclust)
library(e1071)
library(randomForest)
```


```{r, include=FALSE}
setwd("C:/Users/Nana Kwame/Desktop/Multivariate Final Project")
drug <- read.csv("http://archive.ics.uci.edu/ml/machine-learning-databases/00373/drug_consumption.data",header = F)
```

```{r}
colnames(drug) <- c("ID","Age","Gender","Education","Country","Ethnicity","Nscore","Escore","Oscore","Ascore","Cscore","Impulsive","SS","Alcohol","Amphet","Amyl","Benzos","Caff","Cannabis","Choc","Coke","Crack","Ecstasy","Heroine","Ketamine","Legalh","LSD","Meth","Mushrooms","Nicotine","Semer","VSA")
```

#1

##Hotellings For Cannabis
Hotelling's T2 test for the means of NEO-FFI-R scores of Users of cannabis and Non-users of Cannabis.

$H_0:\mu_{Non-users}=\mu_{Users}$ vs $H_a:\mu_{Non-users}\neq\mu_{Users}$

```{r}


group1 <- drug[drug$Cannabis=="CL0"|drug$Cannabis=="CL1",c(7:11)]
group2 <- drug[drug$Cannabis=="CL2"|drug$Cannabis=="CL3"|drug$Cannabis=="CL4"|drug$Cannabis=="CL5"|drug$Cannabis=="CL6",c(7:11)]
HotellingsT2(group1, group2)
```


Here, I conduct a Hotellings Test of equality of mean NEO-FFI-R scores between Users and Non-Users of Cannabis.
Based on the p-value(<0.001), I reject the null hypothesis.
There is convincing evidence that the mean NEO-FFI-R scores of cannabis Users is different from the mean NEO-FFI-R  scores of Non-users of cannabis.(sig.level=0.05)

##Hotellings For Heroine
Hotelling's T2 test for the means of NEO-FFI-R scores of Users of heroine and Non-users of heroine.

$H_0:\mu_{Non-users}=\mu_{Users}$ vs $H_a:\mu_{Non-users}\neq\mu_{Users}$

```{r}
group3 <- drug[drug$Heroine=="CL0"|drug$Heroine=="CL1",c(7:11)]
group4 <- drug[drug$Heroine=="CL2"|drug$Heroine=="CL3"|drug$Heroine=="CL4"|drug$Heroine=="CL5"|drug$Heroine=="CL6",c(7:11)]
HotellingsT2(group3, group4)
```


Here, I conduct a Hotellings Test of equality of mean NEO-FFI-R scores between Users and Non-Users of Heroine.

Based on the p-value(<0.001), I reject the null hypothesis.
There is convincing evidence that the mean NEO-FFI-R scores of Heroine Users is different from the mean NEO-FFI-R  scores of Non-users of Heroine.(sig.level=0.05)

##MANOVA of NEO-FFI-R scores for Users of Cannabis of different Education Levels

H0: Means are the same for all Education levels
HA: At least one mean differ from the others.

```{r}
data_manova <- drug[drug$Cannabis=="CL2"|drug$Cannabis=="CL3"|drug$Cannabis=="CL4"|drug$Cannabis=="CL5"|drug$Cannabis=="CL6",c(7:11,4)]
mn <- manova(as.matrix(data_manova[,1:5])~Education,data=data_manova)
summary(mn)
```

Based on the p-value(<0.001), I reject the null hypothesis at a significance level of 0.05.

There is convincing evidence that the mean NEO-FFI-R scores of cannabis Users is different among users of different Education backgrounds or levels(sig.level=0.05).


##MANOVA of NEO-FFI-R scores for Users of Heroine of different Education Levels

H0: Means are the same for all Education levels
HA: At least one mean differ from the others.

```{r}
dat_manova <- drug[drug$Heroine=="CL2"|drug$Heroine=="CL3"|drug$Heroine=="CL4"|drug$Heroine=="CL5"|drug$Heroine=="CL6",c(7:11,4)]
mn1 <- manova(as.matrix(dat_manova[,1:5])~Education,data=dat_manova)
summary(mn1)
```

Based on the p-value(0.8775), I fail to reject the null hypothesis at a significance level of 0.05.

There is no evidence that the mean NEO-FFI-R scores of Heroine Users is different among users of different Education backgrounds or levels(sig.level=0.05).

#2

##MANOVA of Psychological Variables among Age groups for Cannabis Users

H0: Means are the same for all Age groups
HA: At least one mean differ from the others.

```{r}
data_man <- drug[drug$Cannabis=="CL2"|drug$Cannabis=="CL3"|drug$Cannabis=="CL4"|drug$Cannabis=="CL5"|drug$Cannabis=="CL6",c(7:13,2)]
mn2 <- manova(as.matrix(data_man[,1:7])~Age,data=data_man)
summary(mn2)
```

Based on the p-value(<0.001), I reject the null hypothesis at a significance level of 0.05.

There is convincing evidence that the mean psychological variables scores of cannabis Users is different among different age groups of users(sig.level=0.05).


##MANOVA of Psychological scores for Users of Heroine of different Age groups

H0: Means are the same for all groups
HA: At least one mean differ from the others.

```{r}
dat_m <- drug[drug$Heroine=="CL2"|drug$Heroine=="CL3"|drug$Heroine=="CL4"|drug$Heroine=="CL5"|drug$Heroine=="CL6",c(7:13,2)]
mn2 <- manova(as.matrix(dat_m[,1:7])~Age,data=dat_m)
summary(mn2)
```

Based on the p-value(0.1967), I fail to reject the null hypothesis at a significance level of 0.05.

There is no evidence that the mean psychological scores of Heroine Users is different among users of different Age groups(sig.level=0.05).







#3 
##PCA

```{r}
drug_pca <- drug[,2:13]

cor(drug_pca)

pca1 <- prcomp(drug_pca, scale. = T, retx=T)
plot(pca1)
#biplot(pca1)
summary(pca1)
plot(pca1$sdev,main = "scree plot")
plot(cumsum(pca1$sdev)/sum(pca1$sdev),main = "Cumulative Scree Plot")

pca1$rotation
plot(pca1$rotation[,1],main = "Loadings Plot of PC 1")
plot(pca1$rotation[,2],main = "Loadings Plot of PC 2")
plot(pca1$rotation[,3],main = "Loadings Plot of PC 3")
# biplot(pca1)
# abline(h=0)
```
Since the variables are of different scales, I used the correlation matrix to in computing the principal components. I have decided to go with the first 3 principal components as a way of dimemsionality reduction since, together they explain more than 50% of the total variation as indicated by the screeplot. It is however worthy of mention that the first 3 principal components explain significantly higher amount (about 50% of the total variation) of the total variation than each of the other principal components (i.e using the elbow rule).

  * From the loadings plot of principal component 1, it can be observed that the variables SS and Impulse contributed most in terms of weight.Thus individuals with a high component score will be those with high SS and Impulsive score as well as of younger age.
  * The variables Oscore and Escore contributed most to the second principal component.Thus individuals with a high component score will be those with high NScore and low Oscore and Escore.
  * The variables Education and Nscore contributed most to the Third principal component.Thus individuals with a high component score will be those with high scores in Education and NScore.  


It is worth noting that the original variables are not strongly correlated except for the correlation between the SS and Impulsive score variables. PCA pewrforms better when the original variables are highly correlated.


#4
##Factor Analysis


```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(psych)

x <- scale(drug_pca)
cor.x <- cor(x[,2:9])
p <- dim(x)[2]-8

for (i in 1:p){
  p.vals <- fa(cor.x, nfactors=i, rotate="varimax", scores="regression",n.obs = dim(x)[1], fm="ml")$PVAL 
  print(p.vals)
  }

fa.ml <- factanal(x,factors=4, rotation="varimax",fm="ml")
fa.ml

load <- fa.ml$loadings[,1:4]
for(i in 1:4){ plot(load[,i],main = paste("Loadings plot for Factor",i))}
#plot(load,type="n") 
#text(load,labels=names(x),cex=.7)
```


Looking at the p-values, it was best to do factor analysis with 4 factors.

The Nscore and Escore are highly influenced or driven by the factor 1.We may name factor 1 the "Nscore-Escore Factor".The variables Country and Oscore are highly influenced by Factor 2. We may as well call it the "Country-Oscore factor".The factor 3 is common to Impulsive and SS scores. It can be called the "Risk Factor" according to my understanding of the meanings of the variables.Factor 4 seem to influence all variables evenly but especially Gender, Education and Country.

 
#6.
##Cluster Analysis

```{r,include=FALSE}


regroup<-mutate(drug[,2:32], Alcohol2=ifelse(Alcohol=="CL0"|Alcohol=="CL1",1,2),
                      Amphet2=ifelse(Amphet=="CL0"|Amphet=="CL1",1,2),
                      Amyl2=ifelse(Amyl=="CL0"|Amyl=="CL1",1,2),
                      Benzos2=ifelse(Benzos=="CL0"|Benzos=="CL1",1,2),
                      Caff2=ifelse(Caff=="CL0"|Caff=="CL1",1,2),
                      Cannabis2=ifelse(Cannabis=="CL0"|Cannabis=="CL1",1,2),
                      Choc2=ifelse(Choc=="CL0"|Choc=="CL1",1,2),
                      Coke2=ifelse(Coke=="CL0"|Coke=="CL1",1,2),
                      Crack2=ifelse(Crack=="CL0"|Crack=="CL1",1,2),
                      Ecstasy2=ifelse(Ecstasy=="CL0"|Ecstasy=="CL1",1,2),
                      Heroine2=ifelse(Heroine=="CL0"|Heroine=="CL1",1,2),
                      Ketamine2=ifelse(Ketamine=="CL0"|Ketamine=="CL1",1,2),
                      Legalh2=ifelse(Legalh=="CL0"|Legalh=="CL1",1,2),
                      LSD2=ifelse(LSD=="CL0"|LSD=="CL1",1,2),
                      Meth2=ifelse(Meth=="CL0"|Meth=="CL1",1,2),
                      Mushrooms2=ifelse(Mushrooms=="CL0"|Mushrooms=="CL1",1,2),
                      Nicotine2=ifelse(Nicotine=="CL0"|Nicotine=="CL1",1,2),
                      Semer2=ifelse(Semer=="CL0"|Semer=="CL1",1,2),
                      VSA2=ifelse(VSA=="CL0"|VSA=="CL1",1,2))

```


##K-medoids Clustering

```{r}
x <- scale(drug[,2:13])
#str(drug)
n <- dim(x)[1]
w <- as.matrix(dist(x))  # Generating the distance matrix



## Cluster silhouette
avg.sil <- rep(0, 10)
for(m in 2:9) {
	cl <- pam(x, k=m)  # k-medoids clustering
	sil <- silhouette(cl$clustering, dist(x))
	avg.sil[m] <- mean(sil[,3])  	##Finding average silhouette for the clustering
}
avg.sil

##perform Partitioning Around Medoids
cl<-pam(x,k=2)
cl.table<-table(cl$clustering)
# cl.table<-t(as.matrix(cl.table))
# colnames(cl.table)<-c("cluster1","cluster2")
cl.table

table(cl$clustering,regroup$Alcohol2)

table(cl$clustering,regroup$Amphet2)

table(cl$clustering,regroup$Amyl2)

table(cl$clustering,regroup$Benzos2)

table(cl$clustering,regroup$Cannabis2)

table(cl$clustering,regroup$Choc2)

table(cl$clustering,regroup$Coke2)

table(cl$clustering,regroup$Crack2)

table(cl$clustering,regroup$Ecstasy2)

table(cl$clustering,regroup$Heroine2)

table(cl$clustering,regroup$Ketamine2)

table(cl$clustering,regroup$Legalh2)  ##best clustering fit

table(cl$clustering,regroup$Meth2)

table(cl$clustering,regroup$Mushrooms2)

table(cl$clustering,regroup$Nicotine2)

table(cl$clustering,regroup$Semer2)

table(cl$clustering,regroup$VSA2)

```

The clustering algorithm gave a fairly good fit for a few drugs like the legalh, Ecstacy and Amphetamine. For these drugs there were fewer misclassifications relative to the other drugs. Misclassification errors were very high in Alcohol, chocolate, heroine and crack.
 
 

## Heirarchical Clustering for Cannabis

```{r}
cannabis_data <- drug[,c(2:13,19)]
cannabis_data[,14] <- NA
cannabis_data$V14[which(cannabis_data$Cannabis=="CL0"|cannabis_data$Cannabis=="CL1")]<-"Non-User"
cannabis_data$V14[which(cannabis_data$Cannabis=="CL2"|cannabis_data$Cannabis=="CL3"|cannabis_data$Cannabis=="CL4"|cannabis_data$Cannabis=="CL5"|cannabis_data$Cannabis=="CL6"|cannabis_data$Cannabis=="CL7")]<-"User"
cannabis_data <- cannabis_data[,-13]
names(cannabis_data)[names(cannabis_data) == "V14"] <- "Cannabis"
cannabis_data$Cannabis <- as.factor(cannabis_data$Cannabis)

x <- cannabis_data[,1:12]
hc <- hclust(dist(x), method="complete")
cl <- cutree(hc, k=2)
table(cannabis_data$Cannabis)
table(cannabis_data$Cannabis,cl)
```


## Heirarchical Clustering for Heroine

```{r}
heroine <- drug[,c(2:13,24)]
heroine[,14] <- NA
heroine$V14[which(heroine$Heroine=="CL0"|heroine$Heroine=="CL1")]<-"Non-User"
heroine$V14[which(heroine$Heroine=="CL2"|heroine$Heroine=="CL3"|heroine$Heroine=="CL4"|heroine$Heroine=="CL5"|heroine$Heroine=="CL6"|heroine$Heroine=="CL7")]<-"User"
heroine <- heroine[,-13]
names(heroine)[names(heroine) == "V14"] <- "Heroine"
heroine$Heroine <- as.factor(heroine$Heroine)

x <- heroine[,1:12]
hc <- hclust(dist(x), method="complete")
cl <- cutree(hc, k=2)
table(heroine$Heroine)
table(heroine$Heroine,cl)
```

It is observed that there is a high level of misclassification using the heirarchical clustering with complete linkage.

 
#7
##Classification
 
###Classification for Cannabis

```{r ,message=FALSE,warning=FALSE}
cannabis_data <- drug[,c(2:13,19)]
cannabis_data[,14] <- NA
cannabis_data$V14[which(cannabis_data$Cannabis=="CL0"|cannabis_data$Cannabis=="CL1")]<-"Non-User"
cannabis_data$V14[which(cannabis_data$Cannabis=="CL2"|cannabis_data$Cannabis=="CL3"|cannabis_data$Cannabis=="CL4"|cannabis_data$Cannabis=="CL5"|cannabis_data$Cannabis=="CL6"|cannabis_data$Cannabis=="CL7")]<-"User"
cannabis_data <- cannabis_data[,-13]
names(cannabis_data)[names(cannabis_data) == "V14"] <- "Cannabis"
cannabis_data$Cannabis <- as.factor(cannabis_data$Cannabis)

train_cannabis <- cannabis_data[1:1500,] 
test_cannabis <- cannabis_data[1501:1885,]

##random forest
set.seed(9999)
mod.cannabis<-randomForest(Cannabis ~., data = train_cannabis, ntree = 1000,importance=T)

prediction<-predict(mod.cannabis,test_cannabis)

table(test_cannabis$Cannabis)
table(test_cannabis$Cannabis,prediction)
```

There are a few misclassifications. 47 non-users have been wrongly classified as Cannabis Users and 21 users of cannabis have been classified as Non-users.

###Classification for Alcohol

```{r ,message=FALSE,warning=FALSE}
alcohol_data <- drug[,c(2:13,14)]
alcohol_data[,14] <- NA
alcohol_data$V14[which(alcohol_data$Alcohol=="CL0"|alcohol_data$Alcohol=="CL1")]<-"Non-User"
alcohol_data$V14[which(alcohol_data$Alcohol=="CL2"|alcohol_data$Alcohol=="CL3"|alcohol_data$Alcohol=="CL4"|alcohol_data$Alcohol=="CL5"|alcohol_data$Alcohol=="CL6"|alcohol_data$Alcohol=="CL7")]<-"User"
alcohol_data <- alcohol_data[,-13]
names(alcohol_data)[names(alcohol_data) == "V14"] <- "Alcohol"
alcohol_data$Alcohol <- as.factor(alcohol_data$Alcohol)

train_alcohol <- alcohol_data[1:1500,] 
test_alcohol <- alcohol_data[1501:1885,]

##random forest
set.seed(9999)
mod.alcohol<-randomForest(Alcohol ~., data = train_alcohol, ntree = 1000,importance=T)

prediction<-predict(mod.alcohol,test_alcohol)

table(test_alcohol$Alcohol)
table(test_alcohol$Alcohol,prediction)
```
All 11 non-users in the test data have been incorrectly classified as users of alcohol. All users of alcohol are correctly classified as users of alcohol.







